Estimate of the Neural Net Dimension Using Algebraic Topology and Lie Theory\thanks{The code can be found at: \href{https://github.com/karhunenloeve/ neural-topology-lie}{https://github.com/karhunenloeve/neural-topology-lie}.}

### 2020-07-14 llncs2e/paper12.tex
corrections

\author{Luciano Melodia\orcidID{0000-0002-7584-7287} \\
\and Richard Lenz\orcidID{0000-0003-1551-4824}}
%
\authorrunning{L. Melodia et al.}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{Chair of Computer Science 6\\
Friedrich-Alexander University Erlangen-Nürnberg\\
91058 Erlangen, Deutschland \\
\email{\{luciano.melodia,richard.lenz\}@fau.de}}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
In this paper we present an approach to determine the smallest possible number of perceptrons in a neural net in such a way that the topology of the input space can be learned sufficiently well. We introduce a general procedure based on persistent homology to investigate topological invariants of the manifold on which we suspect the data set. We specify the required dimensions precisely, assuming that there is a smooth manifold on or near which the data are located. Furthermore, we require that this space is connected and has a commutative group structure in the mathematical sense. These assumptions allow us to derive a decomposition of the underlying space whose topology is well known. We use the representatives of the $k$-dimensional homology groups from the persistence landscape to determine an integer dimension for this decomposition. This number is the dimension of the embedding that is capable of capturing the topology of the data manifold. We derive the theory and validate it experimentally on toy data sets.

@@

\section{Introduction}
Since the development of deep neural nets, their parameterization, in particular the smallest possible number of perceptual units in a layer, has been studied. This number is of importance for auto-encoding tasks that require extrapolation or interpolation of data, such as blind source separation or super-resolution. If this number is overestimated, unnecessary resources are consumed during training. If the number is too small, no sufficiently good estimate can be given. To solve this problem, we make the following contribution in this paper:
\begin{itemize}
    \item We study topological invariants using statistical summaries of persistent homology on a filtered simplicial complex on the data.
    \item Using the theory of Lie groups, we specify the smallest possible embedding dimension so that the neural net is able to estimate a projection onto a space with the determined invariants.
\end{itemize}

\section{Related work}
We present chronologically relevant work on neural nets from a differential geometric perspective. Further, we summarize earlier attempts to determine depth and width of a neural net, depending on the data to be processed.

\subsubsection*{Differential geometry in neural nets} The manifold of a dense neural net is in most cases Euclidean. Recently there have been results in the direction of other manifolds on which neural nets can operate. The spherical neural nets operate on a manifold with a commutative group structure \cite{CohenGKW18}.

@@

\subsubsection*{Embedding dimension of neural nets} Cybenko showed that dense neural nets with one hidden layer can approximate any continuous function with a bounded domain with arbitrarily small error \cite{Cybenko92}. The upper bound of the embedding dimension for ReLU nets was recently determined to be $(n+4)$, where $n$ is the sample dimension \cite{RaghuPKGS17}. Bartlett et al. show an almost tight bound for VC dimensions. A neural net with $W$ weights and $L$ layers has $O(WL \log W)$ but $\Omega(WL \log W/L)$ VC dimensions \cite{BartlettHLM19}. Its depth was determined as a parameter depending on the moduli of continuity of the function to be approximated \cite{LinJ18}.

To these fundamental results we contribute a topological approach to parameterization. Similar to Futagami et al.\cite{FutagamiYS19}, we investigate the manifold on which we suspect the data. Our approach is applicable to topological spaces in general and is based on persistent homology. We look for the dimension of a space with the same topological properties that the data indicates and which can be approximated. We assume a simple decomposition into a product space of realaxes and $1$-spheres to get the same homology groups as the persistent ones on a filtration of data. The dimension of the decomposition is a computable lower bound.

\section{Smooth manifolds, Lie groups and persistent homology}
The assumption that a set of points lies on a manifold is more accurate than the treatment in Euclidean space. A \emph{topological manifold} $M$ is a \emph{Hausdorff} space, which means that any two points $x,y \in M$ have always open neighborhoods $U_x,U_y \subset M$, so that their intersection is empty. This behaviour gives an intuitive feeling for the position of points in space. The topology $(M, \tau)$ is a set system that describes the structure of a geometric object. Here $M$ itself and $\emptyset$ must be contained in $\tau$, and any union of open sets from $\tau$ and any intersection of $\tau$ must be contained in $\tau$. If this set system has a subset which generates the topology $\tau$ with finitely many unions, then this subset is a \emph{basis} of the topology. We demand of a manifold that the basis is at most countable. In addition, $M$ is locally Euclidean, i.e. each point has a neighborhood which can be mapped homeomorphically to a subset of $\mathbb{R}^n$. The integer $n$ is the dimension of $M$.

@@

\subsubsection{Lie groups} The study of group theory deals with symmetries which can be expressed algebraically. A pair $(M,\bullet)$, consisting of a map $\bullet: M\times M \rightarrow M$ is called group if the map is associative, $x \bullet (y \bullet z) = (x \bullet y) \bullet z$, has a neutral element so that $x \bullet e_{M} = x$ and has an inverse element to each element of the group, $x \bullet x^{-1} = e_{M}$. A Lie group is a smooth manifold which is equipped with a group structure such that the maps $\bullet:M \times  M \rightarrow  M, (x,y) \mapsto xy, $ and $\imath: M \rightarrow  M, x \mapsto x^{-1}$ are smooth. We call a space \emph{connected} if it cannot be divided into disjoint open neighbourhoods. A group is called Abelian or commutative when all elements under the group operation commute.

@@

In practice neural nets are used with a different number of perceptrons per layer. However, a layer can represent the manifold of the previous one. Manifolds can always be immersed and submersed as long as the rank of the Jacobian of the respective map does not change. Thus, the dimension of the data manifold is equal to the width of the smallest possible layer of a neural net \cite{HauserR17}.

>>>>

\author{Luciano Melodia\orcidID{0000-0002-7584-7287} \\
\and Richard Lenz\orcidID{0000-0003-1551-4824}}
%
\authorrunning{L. Melodia et al.}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{Chair of Computer Science 6\\
Friedrich-Alexander University Erlangen-Nürnberg\\
91058 Erlangen, Deutschland \\
\email{\{luciano.melodia,richard.lenz\}@fau.de}}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
In this paper we present an approach to determine the smallest possible number of neurons in a neural net in such a way that the topology of the input space can be learned sufficiently well. We introduce a general procedure based on persistent homology to investigate topological invariants of the manifold on which we suspect the data set. We specify the required dimensions precisely, assuming that there is a smooth manifold on or near which the data are located. Furthermore, we require that this space is connected and has a commutative group structure in the mathematical sense. These assumptions allow us to derive a decomposition of the underlying space whose topology is well known. We use the representatives of the $k$-dimensional homology groups from the persistence landscape to determine an integer dimension for this decomposition. This number is the dimension of the embedding that is capable of capturing the topology of the data manifold. We derive the theory and validate it experimentally on toy data sets.

@@

\section{Introduction}
Since the development of deep neural nets, their parameterization, in particular the smallest possible number of neurons in a layer, has been studied. This number is of importance for auto-encoding tasks that require extrapolation or interpolation of data, such as blind source separation or super-resolution. If this number is overestimated, unnecessary resources are consumed during training. If the number is too small, no sufficiently good estimate can be given. To solve this problem, we make the following contribution in this paper:
\begin{itemize}
    \item We study topological invariants using statistical summaries of persistent homology on a filtered simplicial complex on the data.
    \item Using the theory of Lie groups, we specify the smallest possible embedding dimension so that the neural net is able to estimate a projection onto a space with the determined invariants.
\end{itemize}

\section{Related work}
We present relevant work on neural nets from a differential geometric perspective. Further, we summarize earlier attempts to determine depth and width of a neural net, depending on the data to be processed.

\subsubsection*{Differential geometry in neural nets} The manifold of a dense neural net is in most cases Euclidean. Recently there have been results in the direction of other manifolds on which neural nets can operate, such as the spherical neural nets \cite{CohenGKW18}. They operate on a manifold with a commutative group structure.

@@

\subsubsection*{Embedding dimension of neural nets} Cybenko showed that dense neural nets with one hidden layer can approximate any continuous function with a bounded domain with arbitrarily small error \cite{Cybenko92}. The upper bound of the embedding dimension for ReLU nets was recently determined to be $(n+4)$, where $n$ is the sample dimension \cite{RaghuPKGS17}. Bartlett et al. show an almost tight bound for VC dimensions. A neural net with $W$ weights and $L$ layers is bounded by $O(WL \log W)$ VC dimensions but has $\Omega(WL \log W/L)$ VC dimensions \cite{BartlettHLM19}. Its depth was determined as a parameter depending on the moduli of continuity of the function to be approximated \cite{LinJ18}.

To these fundamental results we contribute a topological approach to the parameterization of the minimal amount of neurons within a layer, such that the topology of the data manifold can be represented. Similar to Futagami et al.\cite{FutagamiYS19}, we investigate the manifold on which we suspect the data. Our approach is applicable to topological spaces in general and is based on persistent homology. We look for the dimension of a space with the same topological properties that the data indicates and which can be approximated. We assume a simple decomposition into a product space of realaxes and $1$-spheres to get the same homology groups as the persistent ones on a filtration of data. The dimension of the decomposition is a computable lower bound.

\section{Smooth manifolds, Lie groups and persistent homology}
The assumption that a set of points lies on a manifold is more accurate than the treatment in Euclidean space. A \emph{topological manifold} $M$ is a \emph{Hausdorff} space, which means that any two points $x,y \in M$ have always open neighborhoods $U_x,U_y \subset M$, so that their intersection is empty. This behaviour gives an intuitive feeling for the position of points in space. The topology $(M, \tau)$ is a set system that describes the structure of a geometric object. Here $M$ itself and the empty set $\emptyset$ must be contained in $\tau$, and any union of open sets from $\tau$ and any intersection of $\tau$ must be contained in $\tau$. If this set system has a subset which generates the topology $\tau$ with finitely many unions, then this subset is a \emph{basis} of the topology. We demand of a manifold that the basis is at most countable. In addition, $M$ is locally Euclidean, i.e. each point has a neighborhood which can be mapped homeomorphically to a subset of $\mathbb{R}^n$. The integer $n$ is the dimension of $M$.

@@

\subsubsection{Lie groups} The study of group theory deals with symmetries which can be expressed algebraically. A pair $(M,\bullet)$, consisting of a map $\bullet: M\times M \rightarrow M$ is called group if the map is associative, i.e. $x \bullet (y \bullet z) = (x \bullet y) \bullet z$ and has a neutral element so that $x \bullet e_{M} = x$. We also require an inverse element to each element of the group, i.e. $x \bullet x^{-1} = e_{M}$. A Lie group is a smooth manifold which is equipped with a group structure such that the maps $\bullet:M \times  M \rightarrow  M, (x,y) \mapsto xy, $ and $\imath: M \rightarrow  M, x \mapsto x^{-1}$ are smooth. We call a space \emph{connected} if it cannot be divided into disjoint open neighbourhoods. A group is called Abelian or commutative when all elements under the group operation commute.

@@

In practice neural nets are used with a different number of neurons per layer. However, a layer can represent the manifold of the previous one. Manifolds can always be immersed and submersed as long as the rank of the Jacobian of the respective map does not change. Thus, the dimension of the data manifold is equal to the width of the smallest possible layer of a neural net \cite{HauserR17}.
